---
title: "R and Python in one document"
author: "Harshil Sharma"
date: 2026-09-25
description: "A bonus post: R and Python share one session via reticulate, passing objects between them in both directions."
---

Data: [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/), Palmer Station
Antarctica LTER, released under a [CC0 1.0 licence](https://allisonhorst.github.io/palmerpenguins/LICENSE.html).
Citation: Horst AM, Hill AP, Gorman KB (2020). palmerpenguins R package version 0.1.1.
Loaded here through the Python package, again with no network access needed.

The [R post](../penguin-body-mass/index.qmd) and the
[Python post](../penguin-species-classifier/index.qmd) each stay in one language. This one
doesn't: it's a single `.qmd` where `{r}` chunks run through `knitr`, and `knitr` hands
`{python}` chunks to the same Python session via
[reticulate](https://rstudio.github.io/reticulate/). An object computed in one language is
read directly by the other, no file written to disk in between.

## Pointing reticulate at this project's environment

`renv` manages this project's R packages, but reticulate still needs to be told which
Python to use. It should be the same `.venv` that `uv run quarto render` builds two
directories up from this post.

```{r}
#| label: setup-reticulate
#| message: false
#| warning: false

library(reticulate)

venv_path <- normalizePath(file.path(getwd(), "..", "..", ".venv"), mustWork = TRUE)
use_virtualenv(venv_path, required = TRUE)
py_config()$python
```

## Python computes, R plots

A Python chunk loads the penguins and works out the mean flipper length per species.

```{python}
#| label: python-compute

from palmerpenguins import load_penguins

penguins = load_penguins()
species_summary = (
    penguins.dropna(subset=["flipper_length_mm"])
    .groupby("species", as_index=False)["flipper_length_mm"]
    .mean()
    .rename(columns={"flipper_length_mm": "mean_flipper_mm"})
)
species_summary
```

`py$species_summary` reads that pandas `DataFrame` straight into R as a data frame, with no
CSV round-trip, and ggplot draws it.

```{r}
#| label: fig-flipper-by-species
#| fig-cap: "Mean flipper length by species: computed in Python, plotted in R."
#| message: false

library(ggplot2)

species_summary_r <- py$species_summary

ggplot(species_summary_r, aes(x = species, y = mean_flipper_mm, fill = species)) +
  geom_col(show.legend = FALSE) +
  labs(x = "Species", y = "Mean flipper length (mm)") +
  theme_minimal(base_size = 13)
```

## R computes, Python uses it back

The direction reverses just as easily. An R chunk sets a threshold, and a Python chunk
reads it straight out of the R session with `r.`.

```{r}
#| label: r-threshold

heavy_threshold_g <- 4500
heavy_threshold_g
```

```{python}
#| label: python-use-r-object

heavy_count = int((penguins["body_mass_g"] > r.heavy_threshold_g).sum())
total = len(penguins.dropna(subset=["body_mass_g"]))
print(f"{heavy_count} of {total} penguins weigh more than {r.heavy_threshold_g:.0f}g")
```

## What this shows

Nothing here needed a shared file or a second render pass: `species_summary` crossed from
Python into R as `py$species_summary`, and `heavy_threshold_g` crossed back from R into
Python as `r.heavy_threshold_g`, in the same `quarto render` call. The only extra setup was
telling reticulate which virtual environment to use, since without that line it falls back
to whatever Python it finds first on the system rather than this project's pinned one.
