---
title: "What predicts a penguin's body mass?"
author: "Harshil Sharma"
date: 2026-09-25
description: "An EDA and linear model of penguin body mass using the Palmer Penguins dataset, in R."
---

Data: [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/), Palmer Station
Antarctica LTER, released under a [CC0 1.0 licence](https://allisonhorst.github.io/palmerpenguins/LICENSE.html).
Citation: Horst AM, Hill AP, Gorman KB (2020). palmerpenguins R package version 0.1.1.
The data ships inside the `palmerpenguins` R package, so no download or network access is
needed to reproduce this post.

Three species of penguin were measured on three islands in the Palmer Archipelago,
Antarctica. I wanted to know: how much of a penguin's body mass can be explained by its
bill and flipper measurements, once species and sex are accounted for?

```{r}
#| label: setup
#| message: false
#| warning: false

library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(broom)
library(knitr)
```

## Cleaning the data

A handful of rows are missing measurements or sex entirely. Since the model below needs
complete cases across all predictors, I drop them here rather than silently letting `lm()`
do it, and report how much data that costs me.

```{r}
#| label: clean-data

penguins_clean <- penguins |>
  filter(!is.na(sex)) |>
  select(species, island, bill_length_mm, bill_depth_mm, flipper_length_mm,
         body_mass_g, sex) |>
  na.omit()

n_dropped <- nrow(penguins) - nrow(penguins_clean)
cat("Rows in raw data:", nrow(penguins), "\n")
cat("Rows after dropping incomplete cases:", nrow(penguins_clean), "\n")
cat("Rows dropped:", n_dropped, "\n")
```

That drops `r n_dropped` of the original `r nrow(penguins)` rows, leaving
`r nrow(penguins_clean)` penguins with complete measurements.

## Body mass by species and sex

```{r}
#| label: summary-table

penguin_summary <- penguins_clean |>
  group_by(species, sex) |>
  summarise(
    n = n(),
    mean_mass_g = round(mean(body_mass_g), 1),
    sd_mass_g = round(sd(body_mass_g), 1),
    mean_flipper_mm = round(mean(flipper_length_mm), 1),
    .groups = "drop"
  )

kable(penguin_summary,
      caption = "Body mass and flipper length by species and sex.")
```

Gentoo penguins are noticeably heavier than Adelie or Chinstrap, and within every species
males run heavier than females. Flipper length tracks the same pattern, which suggests it
might be a useful predictor of mass alongside bill measurements.

```{r}
#| label: fig-penguin-mass
#| fig-cap: "Body mass of penguins by species, split by sex."
#| warning: false

ggplot(penguins_clean, aes(x = species, y = body_mass_g, fill = sex)) +
  geom_boxplot(alpha = 0.8) +
  labs(
    x = "Species",
    y = "Body mass (g)",
    fill = "Sex",
    title = "Gentoo penguins are the heaviest of the three species"
  ) +
  theme_minimal(base_size = 13)
```

## Modelling body mass

I fit a linear model of body mass against bill length, flipper length, species, and sex.
Species and sex are included because @fig-penguin-mass suggests both shift the whole
distribution, not just the average.

```{r}
#| label: fit-model

mass_model <- lm(
  body_mass_g ~ bill_length_mm + flipper_length_mm + species + sex,
  data = penguins_clean
)

kable(tidy(mass_model, conf.int = TRUE),
      digits = 2,
      caption = "Coefficient estimates for the body mass model.")
```

```{r}
#| label: model-fit

glance_stats <- glance(mass_model)
kable(glance_stats[, c("r.squared", "adj.r.squared", "sigma", "df.residual")],
      digits = 3,
      caption = "Overall fit of the body mass model.")
```

The model explains about `r round(glance_stats$r.squared * 100, 1)`% of the variance in
body mass (adjusted R² = `r round(glance_stats$adj.r.squared, 3)`). Flipper length has the
largest standardised effect of the continuous predictors, being Gentoo adds a large positive
shift on top of that, and being male adds a further few hundred grams on average, holding
the other measurements fixed. Bill length alone is a weaker predictor once flipper length
and species are already in the model, which fits the biology: flipper length and species are
closely tied to overall skeletal size, while bill length varies more independently of body
mass within a species.

## What I found

Body size in these penguins is driven mostly by species and flipper length, with sex adding
a consistent but smaller shift. A simple linear model with four predictors already accounts
for the large majority of the variation in body mass, which suggests there isn't much left
on the table for a more complex model to pick up without additional variables (diet, foraging
range, or year, for instance) that aren't in this dataset.
